---
title: "Covid91 vaccine study Final2024"
author: "Elisa Gonzalez"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, covid19 = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
Male = filter(FinalData, sex =="M")
Female = filter(FinalData, sex =="F")
Gay = filter(FinalData, LGBTQ =="gay")
DrugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

##Overall apporach 

1. **Descriptive Analysis**: Summarize data perceptibly then numerically.  
2. **Inferential Analysis**: Perform hypothesis testing:  
   - \(H_0\): The vaccine has no impact on infection rates.  
   - \(H_A\): The vaccine reduces infection rates.  
3. **Effectiveness Calculation**:
   \[ 
   \text{Vaccine Effectiveness (VE)} = 1 -  \frac{\text{Risk}_{\text{vaccine}}}{\text{Risk}_{\text{placebo}}}
   \]
4. **Effectivness Confidence Interval**: low to high

## Males

```{r}
male_summary <- Male %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
male_summary <- male_summary %>%
  mutate(total = rowSums(select(., -treatment)))
male_column_totals <- male_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
male_summary <- bind_rows(male_summary, male_column_totals)
male_summary
```

Here’s the current data for men: There are 18,392 men in total. Of these, 18,037 are considered healthy, while 355 have COVID. To better illustrate this, let's include a bar plot for a clearer visual representation, as the numbers alone don't effectively show the trend.

###Visual results 
```{r}
# Bar plot

# Summarize the data for Males by infection status and treatment
male_summary_data <- Male %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
male_plot_count <- ggplot(male_summary_data, aes(x = infected, y = count, fill = treatment))
male_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Males Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
male_plot_percent <- ggplot(male_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
male_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Males Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The bar plots above show the number of males infected with COVID-19. In both graphs, the results are similar to the placebo group, but visually, males who received the vaccination seem to have lower COVID rates compared to the placebo group. Although there is a noticeable difference, we need to conduct a statistical analysis to determine if this difference is significant.

### Numerical Results
```{r}
table1 <- xtabs(~infected + treatment, data=Male)
rowPerc(table1)
colPerc(table1)
```

The numerical data shows that 73.8% of the infected males received the placebo, while 26.2% received the vaccine. Among the healthy individuals, 49.99% received the vaccine, and 50.01% received the placebo. Overall, 2.82% of males who received the placebo contracted COVID-19, compared to just 1.02% of those who received the vaccine, suggesting that the vaccination reduces infection rates. These findings support the conclusion that the vaccine is effective in reducing COVID-19 in males, which aligns with the trends shown in the bar plots.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
Malefish <- fisher.test(table1)
Malefish
```

Fisher's Exact Test produces an extremely small p-value of 0.00000000000000022, which is the key result to focus on. This suggests that the observed differences in infection rates between the vaccine and placebo groups are highly unlikely to have occurred by chance. Therefore, we reject the null hypothesis, which posits that treatment has no effect on infection rates. The odds ratio of approximately 2.82 indicates that males who received the vaccine had about a 2.8 times higher risk of contracting COVID-19 compared to those who received the placebo. These results imply that the vaccine has minimal impact on preventing COVID-19 in males.

```{r}
# Calculate effectiveness
male_effect <- Malefish$p.value / (Malefish$p.value + 1)
effectiveness_male <- (1 - 1 / Malefish$estimate) * 100
print(effectiveness_male)

# Calculate effectiveness confidence interval
male_low <- (1 - 1 / Malefish$conf.int[1]) * 100
male_hi <- (1 - 1 / Malefish$conf.int[2]) * 100
options(digits = 2)

#print 
male_low
male_hi
```

### `r effectiveness_male`% Effectiveness with Confidence Interval: `r male_low`%  to `r male_hi`%

The COVID-19 vaccine for men demonstrates an efficacy of 64.48%, surpassing the 50% threshold for success. The confidence interval ranges from 54.75% to 72.32%, reinforcing its effectiveness in reducing infection rates among males. Based on the low p-value from Fisher's Exact Test and an odds ratio of 2.82, we reject the null hypothesis (which suggests no effect) and accept the alternative hypothesis (that the vaccine reduces infection rates). This indicates that vaccinated men are nearly 2.8 times more likely to avoid contracting COVID-19 compared to those who received the placebo. The confidence interval further supports the vaccine’s strong protective effect.


## Females

### Summary of Females
```{r}
female_summary <- Female %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
female_summary <- female_summary %>%
  mutate(total = rowSums(select(., -treatment)))
female_column_totals <- female_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
female_summary <- bind_rows(female_summary, female_column_totals)
female_summary
```

Here’s the data for the women so far: There are 18,328 females in total, with 17,963 being healthy and 365 infected with COVID. Since these numbers alone don’t clearly show a trend, let’s add a bar plot for better visual representation.

### Visual Results

```{r}
# Bar plot

# Summarize the data for Females by infection status and treatment
female_summary_data <- Female %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
female_plot_count <- ggplot(female_summary_data, aes(x = infected, y = count, fill = treatment))
female_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Females Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
female_plot_percent <- ggplot(female_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
female_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Females Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The bar plots above show the number of females infected with COVID-19. In both graphs, the results are similar to the placebo group, but visually, females who received the vaccine appear to have lower COVID rates compared to those who received the placebo. While the difference is noticeable, we need to conduct a statistical analysis to determine if it is significant.

### Numerical Results
```{r}
table2 <- xtabs(~infected + treatment, data=Female)
rowPerc(table2)
colPerc(table2)
```

The numerical data shows that 35% of the infected females received the vaccine, while 65% received the placebo. Among the healthy individuals, half received the vaccine and half received the placebo. Overall, 1.4% of females who received the vaccine contracted COVID-19, compared to 2.6% of those who received the placebo, suggesting that the vaccine reduces infection rates. These findings indicate that the vaccine is effective in lowering COVID-19 rates in females, which is consistent with the bar plots illustrating its efficacy.

### Inferential Results
```{r}
chisq.test(table2)
chisqtestGC(table2)
Femalefish <- fisher.test(table2)
Femalefish
```

The main takeaway from Fisher's Exact Test is that the difference in infection rates between the vaccine and placebo groups is highly unlikely to have occurred by chance, with an extremely small p-value of 0.00000001. As a result, we reject the null hypothesis, which suggests the treatment has no impact on infection rates. The odds ratio of approximately 1.9 indicates that women who received the vaccine had nearly twice the risk of contracting COVID-19 compared to those who received the placebo. Therefore, these results suggest that the vaccine is ineffective at preventing COVID-19 in females. 


```{r}
# Calculate effectiveness
female_effect <- Femalefish$p.value / (Femalefish$p.value + 1)
effectiveness_female <- (1 - 1 / Femalefish$estimate) * 100
print(effectiveness_female)

# Calculate effectiveness confidence interval
female_low <- (1 - 1 / Femalefish$conf.int[1]) * 100
female_hi <- (1 - 1 / Femalefish$conf.int[2]) * 100
options(digits = 2)

#print 
female_low
female_hi
```

### `r effectiveness_female`% Effectiveness with Confidence Interval: `r female_low`%  to `r female_hi`%
 
The vaccine's efficacy for females is shown above. A vaccine is considered effective if its efficacy exceeds 50%. However, with a calculated efficacy of 46.13% and a confidence interval ranging from 32.79% to 56.98%, we conclude that the vaccine is ineffective at reducing infection rates in females. The alternative hypothesis suggested that the vaccine reduces infection rates, while the null hypothesis stated it has no effect. Based on the extremely low p-value from Fisher's Exact Test and an odds ratio of 1.9, we reject the null hypothesis and accept the alternative. This suggests that the vaccine may actually increase the risk of infection in females. The odds ratio indicates that women who received the vaccine were nearly 1.9 times more likely to contract COVID-19 compared to those who received the placebo. The confidence interval further supports this finding, suggesting a modest, and likely negative, impact on vaccine effectiveness.
 
## LGBTQ

### Summary of LGBTQ
```{r}
gay_summary <- Gay %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
gay_summary <- gay_summary %>%
  mutate(total = rowSums(select(., -treatment)))
gay_column_totals <- gay_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
gay_summary <- bind_rows(gay_summary, gay_column_totals)
gay_summary
```

Here is the LGBTQ data available so far: There are 3,060 LGBTQ individuals, with 3,000 being healthy and 60 infected with COVID. Since these numbers alone don't clearly show a trend, let's add a bar plot for better visual representation.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Gay by infection status and treatment
gay_summary_data <- Gay %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
gay_plot_count <- ggplot(gay_summary_data, aes(x = infected, y = count, fill = treatment))
gay_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
gay_plot_percent <- ggplot(gay_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
gay_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The bar plots above show the number of LGBTQ individuals infected with COVID-19. Visually, gay people who received the vaccine appear to have lower COVID rates compared to those who received the placebo. While the differences are quite noticeable, a statistical analysis is needed to determine if the difference is statistically significant.

### Numerical Results
```{r}
table3 <- xtabs(~infected + treatment, data=Gay)
rowPerc(table3)
colPerc(table3)
```

The numerical data shows that 20% of the gay individuals who contracted COVID had received the vaccine, while 80% had received the placebo. Among the healthy individuals, 52% received the vaccine and 48% received the placebo. Overall, 0.77% of gay individuals who received the vaccine contracted COVID-19, compared to 3.2% of those who received the placebo, suggesting that the vaccine reduces infection rates. These results indicate that the vaccine is effective in reducing the prevalence of COVID-19 among LGBTQ individuals, which aligns with the trends shown in the bar plots demonstrating its efficacy.

### Inferential Results
```{r}
chisq.test(table3)
chisqtestGC(table3)
Gayfish <- fisher.test(table3)
Gayfish
```

The key finding from Fisher's Exact Test is a very small p-value of 0.0000007, which suggests that the observed differences in infection rates between the vaccine and placebo groups are highly unlikely to have occurred by chance. As a result, we reject the null hypothesis, which posits that treatment has no effect on infection rates. The odds ratio of approximately 4.3 indicates that gay individuals who received the vaccine had about a 4.3 times higher risk of contracting COVID-19 compared to those who received the placebo. These results suggest that the vaccine does not protect gay individuals from COVID-19.

```{r}
# Calculate effectiveness
gay_effect <- Gayfish$p.value / (Gayfish$p.value + 1)
effectiveness_gay <- (1 - 1 / Gayfish$estimate) * 100
print(effectiveness_gay)

# Calculate effectiveness confidence interval
gay_low <- (1 - 1 / Gayfish$conf.int[1]) * 100
gay_hi <- (1 - 1 / Gayfish$conf.int[2]) * 100
options(digits = 2)

#print 
gay_low
gay_hi
```

### `r effectiveness_gay`% Effectiveness with Confidence Interval: `r gay_low`%  to `r gay_hi`%

The efficacy of the COVID-19 vaccine for gay individuals is shown above. A vaccine is considered effective if its efficacy exceeds 50%. With an efficacy of 76.6% and a confidence interval ranging from 55.06% to 88.74%, we can conclude that the vaccine is effective in reducing infection rates among gay individuals. The alternative hypothesis suggests that the vaccine lowers infection rates, while the null hypothesis proposes that it has no effect. The extremely low p-value from Fisher's Exact Test and an odds ratio of 4.3 lead us to reject the null hypothesis and accept the alternative. This indicates that the vaccine is effective in lowering infection rates. However, the odds ratio suggests that gay individuals who received the vaccine had nearly a 4.3 times higher risk of contracting COVID-19 compared to those who received the placebo. The confidence interval further supports this, confirming that the vaccine has a significant protective effect for the LGBTQ community.   

## Druggies

### Summary of DrugUser
```{r}
druguser_summary <- DrugUser %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)

# Add total column
druguser_summary <- druguser_summary %>%
  mutate(total = rowSums(select(., -treatment)))

# Add total row for each treatment
druguser_column_totals <- druguser_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")

# Combine the totals with the main data
druguser_summary <- bind_rows(druguser_summary, druguser_column_totals)

druguser_summary
```

Here is the data provided for drug users so far: There are 3,060 individuals who use drugs, with 3,000 being healthy and 60 infected with COVID. Since these numbers alone don't clearly show a trend, let's add a bar plot for better visual representation.

### Visual Results
```{r}
# Summarize the data for DrugUser by infection status and treatment
druguser_summary_data <- DrugUser %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
druguser_plot_count <- ggplot(druguser_summary_data, aes(x = infected, y = count, fill = treatment))
druguser_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
druguser_plot_percent <- ggplot(druguser_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
druguser_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The bar plots above show the number of drug users infected with COVID-19. Visually, drug users who received the vaccine appear to have higher COVID rates compared to those who received the placebo. Although the difference looks notable, a statistical analysis is needed to assess whether it is statistically significant.

### Numerical Results

```{r}
table4 <- xtabs(~infected + treatment, data=DrugUser)
rowPerc(table4)
colPerc(table4)
```

The numerical data shows that 91.7% of infected drug users had received the vaccine, while 8.3% had received the placebo. Among those who are healthy, 49.0% received the vaccine and 51.0% received the placebo. Overall, 3.6% of drug users who received the vaccine contracted COVID-19, compared to just 0.33% of those who received the placebo, suggesting that the vaccine may be associated with a higher risk of infection in this population. These findings indicate that the vaccine might not be effective in reducing COVID-19 incidence among drug users. This is consistent with the bar plots, which visually show higher infection rates among vaccinated drug users.

### Inferential Results

```{r}
chisq.test(table4)
chisqtestGC(table4)
DrugUserfish <- fisher.test(table4)
DrugUserfish
```

The key finding from Fisher's Exact Test is the extremely small p-value of 0.000000000003, which strongly suggests that the observed differences in infection rates between the vaccine and placebo groups are highly unlikely to have occurred by chance. This leads us to reject the null hypothesis, which assumes that treatment has no effect on infection rates. The odds ratio of approximately 0.087 indicates that drug users who received the vaccine had an 11.5-fold lower risk of contracting COVID-19 compared to those who received the placebo. These results suggest that the vaccine is highly effective in preventing COVID-19 in drug users. This aligns with the statistical findings showing lower infection rates among vaccinated individuals in this group.

```{r}
# Calculate effectiveness
DrugUserfish <- fisher.test(table4)
effectiveness_drug <- (1 - DrugUserfish$estimate) * 100
print(effectiveness_drug)

# Calculate effectiveness confidence level
drug_low <- (1 - DrugUserfish$conf.int[2]) * 100  
drug_hi <- (1 - DrugUserfish$conf.int[1]) * 100  
options(digits = 2)

#print
drug_low
drug_hi
```

### `r effectiveness_drug`% Effectiveness with Confidence Interval: `r drug_low`%  to `r drug_hi`%

The efficacy of the COVID-19 vaccine for drug users is shown above. A vaccine is considered effective if its efficacy is over 50%. With an efficacy of 91.3% and a confidence interval of 78.0% to 97.3%, we can conclude that the vaccine is highly effective at reducing infection rates among drug users. The alternative hypothesis suggests the vaccine lowers infection rates, while the null hypothesis says it has no effect. The very low p-value from Fisher's Exact Test and an odds ratio of 0.087 lead us to reject the null hypothesis. This means the vaccine significantly lowers infection rates for drug users. The odds ratio of 0.087 shows that vaccinated drug users have an 11.5 times lower risk of contracting COVID-19 compared to those who received the placebo. Since the odds ratio is less than 1, the vaccine is highly effective. The confidence interval further supports this, confirming the vaccine’s strong protective effect for drug users.

# Overall Results and Conclusions

## Conclusion for Males
For males, the efficacy of the COVID vaccine is 64.48%, with a confidence interval between 54.75% and 72.32%. This conclusion is well supported by statistical tests: The null hypothesis of no vaccine impact was rejected by Fisher's Exact Test, which revealed a highly significant p-value (\(p = 2.2 \times 10^{-16}\)).Male recipients of the vaccination had a nearly 2.8-fold lower risk of infection than those receiving a placebo, according to the odds ratio of 2.82. Overall, the vaccination greatly lowers the risk of illness in males, demonstrating its efficacy in this population.     


